---
title: "Machine Learning Project"
author: "Jason Merefield"
date: "Thursday, April 23, 2015"
output: html_document
---

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset). 

# Goal

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. I will also use the chosen prediction model to predict 20 different test cases.

# Loading up the data

```{r,echo=FALSE}
# load libraries
loadLibraries <- function(){
        require(rpart, quietly = T, warn.conflicts = F)
        require(randomForest, quietly = T, warn.conflicts = F)
        require(caret, quietly = T, warn.conflicts = F)
}

suppressPackageStartupMessages(loadLibraries())
```

```{r}
# load training data
pml.training <- read.csv("~/Coursera/Machine Learning/pml-training.csv", stringsAsFactors=F, na.strings=c("NA","#DIV/0!"))
```
# Transforming the data

The data has a few columns that aren't of interest, and quite a few with all or nealy-all 'NA' values. The following code removes those columns.
The target, column 'classe', is also converted to a factor.

```{r}
# convert target to factor
pml.training$classe <- as.factor(pml.training$classe)
# remove ids and other metadata-type columns from data frame
df <- pml.training[,-c(1:6)]
# remove any columns with high proportion of NAs
t <- sapply(df, function(x) ifelse(sum(is.na(x)) / nrow(df) > 0.5, FALSE, TRUE))
t_df <- df[,t]
# get predictor columns out
cols <- t_df[,-54]
# fix na's up with imputation
imputeModel <- preProcess (cols, method = c("knnImpute"))
cols_final <- predict(imputeModel,newdata = cols)
# put back together the data frame
df_final <- as.data.frame(cbind(cols_final,classe = df$classe))
# set psuedo-random seed to allow reproducability
set.seed(12345)
# split training data
ids <- createDataPartition(y = df_final$classe, p=0.8, list=F)
train_df <- df_final[ids,]
test_df <- df_final[-ids,]
```

# Test prediction accuracy with model 'rpart' (Recursive Partitioning)

First up I used the 'rpart' model to train the model.
For this, and subsequent training, I am using a repeated cross-validation with 3 repeats of 4 folds.

```{r}
# setup training control for caret
fitControl <- trainControl(method="repeatedcv",number=4,repeats=3)
# try 'rpart' model first
rpart_fit <- train(classe ~ ., data=train_df, trControl=fitControl, method='rpart')
# get predictions
train_pred_rpart <- predict(rpart_fit,train_df)
test_pred_rpart <- predict(rpart_fit,test_df)
# get out confusion matrix for each data set
train_cm_rpart <- caret::confusionMatrix(table(train_pred_rpart, train_df$classe))
test_cm_rpart <- caret::confusionMatrix(table(test_pred_rpart, test_df$classe))
# calculate sample errors using the reciprocal of the accuracy
in_sample_error_rpart = 1 - train_cm_rpart$overall[1]
out_of_sample_error_rpart = 1 - test_cm_rpart$overall[1]
```

The out-of-sample error from the 'rpart' model is `r out_of_sample_error_rpart`.

# Test prediction accuracy with model 'rf' (Random Forest)

Next up I tried the old faithful Random Forest model.
I use this as my benchmark in all my Kaggle competitions and have faith it will give me one of the lowest out-of-sample errors.

```{r}
# try 'rf' model next
rf_fit <- train(classe ~ ., data=train_df, trControl=fitControl, method='rf')

# get predictions
train_pred_rf <- predict(rf_fit,train_df)
test_pred_rf <- predict(rf_fit,test_df)

# get out confusion matrix for each data set
train_cm_rf <- caret::confusionMatrix(table(train_pred_rf, train_df$classe))
test_cm_rf <- caret::confusionMatrix(table(test_pred_rf, test_df$classe))

# calculate sample errors using the reciprocal of the accuracy
in_sample_error_rf = 1 - train_cm_rf$overall[1]
out_of_sample_error_rf = 1 - test_cm_rf$overall[1]
```

The out-of-sample error from the 'rpart' model is `r out_of_sample_error_rf`.

# Validation of model

To truly test the model it is advisable to test it against a 'validation' test set; a data set that has not been touched by training and testing.

The lecturer has provided just such a set, of which I ran the RF model through as follows:

```{r}
# load training data
pml.testing <- read.csv("~/Coursera/Machine Learning/pml-testing.csv", stringsAsFactors=F, na.strings=c("NA","#DIV/0!"))
# remove ids and other metadata-type columns from data frame
df <- pml.testing[,-c(1:6,160)]
# remove any columns with high proportion of NAs
t <- sapply(df, function(x) ifelse(sum(is.na(x)) / nrow(df) > 0.5, FALSE, TRUE))
t_df <- df[,t]
# fix na's up with imputation using model created during preprocess of training
cols_final <- predict(imputeModel,newdata = t_df)
# now predict using RF model
new_predictions <- predict(rf_fit, cols_final)
# submit for evaluation
pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("~/Coursera/Machine Learning/For submission/problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}

pml_write_files(new_predictions)
```
